---
title: "Advanced Shiny"
output:
    html_document:
        keep_md: true
---

```{r setup, cache=FALSE, echo=FALSE, message=F, warning=F, tidy=FALSE}
require(knitr)
options(width=100)
opts_chunk$set(message=F, error=F, warning=F, comment=NA, fig.align='center', dpi=100, tidy=F, cache.path='.cache/', fig.path='fig/')

options(xtable.type='html')
knit_hooks$set(inline=function(x) {
    if(is.numeric(x)) {
        round(x, getOptions('digits'))
    } else {
        paste(as.character(x), collapse=', ')
    }
})
knit_hooks$set(plot=knitr:::hook_plot_html)
```

## `shiny` Revisited

- In the last lecture, we covered basic creation of `shiny` applications
- If you tried it and are like most, you had an easy time with `ui.R` but a harder time with `server.R`
- In this lecture, we cover some more of the details of shiny
- Since writing the last lecture, a more detailed tutorial has been created that is worth checking out ([http://shiny.rstudio.com/tutorial/](http://shiny.rstudio.com/tutorial/))

---

## Details

- Code that you put before `shinyServer` in the `server.R` function gets called once when you do `runApp()`
- Code inside the unnamed function of `shinyServer(function(input, output){})` -- but not in a reactive statement -- will run once for every new user (or page refresh)
- Code in reactive functions of `shinyServer` get run repeatedly as needed when new values are entered
    - Reactive functions are those like `render*`

---

## Experiment

### `ui.R`

```{r, eval=FALSE}
library(shiny)
shinyUI(pageWithSidebar(
    headerPanel("Hello Shiny!"),
    sidebarPanel(
        textInput(inputId="text1", label="Input Text1"),
        textInput(inputId="text2", label="Input Text2")
    ),
    mainPanel(
        p('Output text1'),
        textOutput('text1'),
        p('Output text2'),
        textOutput('text2'),
        p('Output text3'),
        textOutput('text3'),
        p('Output text4'),
        textOutput('text4'),
        p('Inside text, but non-reactive'),
        textOutput('text5')
    )
))
```

### 'server.R' -- Set `x <- 0` before running

```{r, eval=FALSE}
library(shiny)
x <<- x + 1
y <<- 0

shinyServer(
    function(input, output) {
        y <<- y + 1
        output$text1 <- renderText({input$text1})
        output$text2 <- renderText({input$text2})
        output$text3 <- renderText({as.numeric(input$text1) + 1})
        output$text4 <- renderText(y)
        output$text5 <- renderText(x)
    }
)
```

---

## Try It

- Type `runApp()`
- Notice hitting refresh increments `y` but entering values in the textbox does not
- Notice `x` is always 1
- Watch how it updated `text1` and `text2` as needed
- Doesn't add 1 to `text1` every time a new `text2` is input
- _Important_ Try `runApp(display.mode='showcase')`

---

## Reactive Expressions

- Sometimes to speed up your app, you want reactive operations (those operations that depend on widget input values) to be performed outside of a `render*` statement
- For example, you want to do some code that gets reused in several `render*` statements and don't want to recalculate it for each
- The `reactive` function is made for this purpose

---

## Example

### `server.R`

```{r, eval=FALSE}
library(shiny)
shinyServer(
    function(input, output) {
        x <- reactive({as.numeric(input$text1) + 100})
        output$text1 <- renderText({x()                          })
        output$text2 <- renderText({x() + as.numeric(input$text2)})
    }
)
```

---

## As opposed to

```{r, eval=FALSE}
library(shiny)
shinyServer(
    function(input, output) {
        output$text1 <- renderText({as.numeric(input$text1) + 100  })
        output$text2 <- renderText({as.numeric(input$text1) + 100 +
            as.numeric(input$text2)})
    }
)
```

---

## Discussion

- Do `runApp(display.mode='showcase')`
- While inconsequential, the second example has to add 100 twice every time text1 is updated for the second set of code
- Also note the somewhat odd syntax for reactive variables

---

## Non-reactive Reactivity (What?)

- Sometimes you don't want `shiny` to immediately perform reactive calculations from widget inputs
- In other words, you want something like a submit button

---

## `ui.R`

```{r, eval=FALSE}
library(shiny)
shinyUI(pageWithSidebar(
    headerPanel("Hello Shiny1"),
    sidebarPanel(
        textInput(inputId="text1", label="Input Text1"),
        textInput(inputId="text2", label="Input Text2"),
        actionButton("goButton", "Go!")
    ),
    mainPanel(
        p('Output text1'),
        textOutput('text1'),
        p('Output text2'),
        textOutput('text2'),
        p('Output text3'),
        textOutput('text3')
    )
))
```

---

## `server.R`

```{r, eval=FALSE}
library(shiny)
shinyServer(
    function(input, output) {
        output$text1 <- renderText({input$text1})
        output$text2 <- renderText({input$text2})
        output$text3 <- renderText({
            input$goButton
            isolate(paste(input$text1, input$text2))
        })
    }
)
```

---

## Try It Out

- Notice it doesn't display output `text3` until the `go` button is pressed
- `input$goButton` (or whatever you named it) gets increased by one for every time pushed
- So, when in reactive code such as `render*` or `reactive`, you can use conditional statements like below to only execute code on the first button press or to not execute until the first or subsequent button press

`if(input$goButton == 1) { Conditional Statements }`

---

## Example

Here's some replaced code from our previous `server.R`

```{r, eval=FALSE}
output$text3 <- renderText({
    if(input$goButton == 0) "You have not pressed the button"
    else if(input$goButton == 1) "You pressed it once"
    else "Okay quit pressing it"
})
```

---

## More On Layouts

- The sidebar layout with a main panel is the easiest
- Using `shinyUI(fluidpage(` is much more flexible and allows tighter access to the bootstrap styles
- Examples here ([http://shiny.rstudio.com/articles/layout-guide.html](http://shiny.rstudio.com/articles/layout-guide.html))
- `fluidRow` statements create rows and then the `column` function from within it can create columns
- Tabsets, navlists, and navbars can be created for more complex apps

---

## Directly Using html

- Fore more complex layouts, direct use of html is preferred ([http://shiny.rstudio.com/articles/html-ui.html](http://shiny.rstudio.com/articles/html-ui.html))
- Also, if you know web development well, you might find using R to create web layouts kind of annoying
- Create a directory called _www_ in the same directory with `server.R`
- Have an `index.html` page in that directory
- Your named input variables will be passed to `server.R`
    - `<input type="number" name="n" value="500" min="1" max="1000" />`
- Your `server.R` output will have class definitions of the form `shiny-*`
    - `<pre id="summary" class="shiny-text-output"></pre>`

---

## Debugging Techniques for `shiny`

- Debugging `shiny` apps can be tricky
- We saw that `runApp(displayMode='showcase')` highlights execution while a `shiny` app runs
- Using `cat` in your code displays output to `stdout` (so R console)
- The `browser()` function can interupt execution and can be called conditionally ([http://shiny.rstudio.com/articles/debugging.html](http://shiny.rstudio.com/articles/debugging.html))