---
title: "R Packages"
output:
    html_document:
        keep_md: true
---

```{r setup, cache=FALSE, echo=FALSE, message=F, warning=F, tidy=FALSE}
require(knitr)
options(width=100)
opts_chunk$set(message=F, error=F, warning=F, comment=NA, fig.align='center', dpi=100, tidy=F, cache.path='.cache/', fig.path='fig/')

options(xtable.type='html')
knit_hooks$set(inline=function(x) {
    if(is.numeric(x)) {
        round(x, getOptions('digits'))
    } else {
        paste(as.character(x), collapse=', ')
    }
})
knit_hooks$set(plot=knitr:::hook_plot_html)
```

## What Is an `R` Package?

- A mechanism for extending the basic functionality of `R`
- A collection of `R` functions, or other (data) objects
- Organized in a systemic fashion to provide a minimal amount of consistency
- Written by users/developers everywhere

---

## Where Are These `R` Packages?

- Primarily availabile from CRAN and Bioconductor
- Also available from GitHub, Bitbucket, Gitorious, etc.
- Packages from CRAN/Bioconductor can be installed with `install.packages()`
- Packages from GitHub can be installed using `install_github()` from the `devtools` package

You do not have to put a package on a central repository, but doing so makes it easier for others to install your package

---

## What's the Point?

- "Why not just make some code available?"
- Documentation / vignettes
- Centralized resources like CRAN
- Minimal standards for reliability and robustness
- Maintainability / extension
- Interface definition / clear API
- Users know that it will at least load properly

---

## Package Development Process

- Write some code in an `R` script file (`.R`)
- Want to make code available to others
- Incorporate `R` script file into `R` package structure
- Write documentation for user functions
- Include some other material (examples, demos, datasets, tutorials)
- Package it up!

---

## Package Development Process

- Submit package to CRAN or Bioconductor
- Push source code repository to GitHub or other source code sharing website
- People find all kinds of problems with your code
    - Scenario #1: They tell you about those problems and expect you to fix it
    - Scenario #2: They fix the problem for you and show you the changes
- You incorporate the changes and release a new version

---

## `R` Package Essentials

- An `R` package is started by creating a directory with the name of the `R` package
- A `DESCRIPTION` file which has info about the package
- `R` code! (in the `R/` subdirectory)
- Documentation (in the `man/` subdirectory)
- `NAMESPACE` (optional, but do it)
- Full requirements in Writing `R` Extensions

---

## The `DESCRIPTION` File

- **Package**: Name of package (e.g. `library(name)`)
- **Title**: Full name of package
- **Description**: Longer description of package in one sentence (usually)
- **Version**: Version number (usually in _M.m-p_ format)
- **Author, Authors**: Name of the original author(s)
- **Maintainer**: Name + email of person who fixes problems
- **License**: License for the source code

---

## The `DESCRIPTION` File

These fields are optional but commonly used

- **Depends**: `R` packages that your package depends on
- **Suggests**: Optional `R` packages that users may want to have installed
- **Date**: Release date in _YYYY-MM-DD_ format
- **URL**: Package home page
- **Other** fields can be added

---

## `DESCRIPTION` File: `gpclib`

**Package**: gpclib

**Title**: General Polygon Clipping Library for R

**Description**: General polygon clipping routines for R based on Alan Murta's C library

**Version**: 1.5-5

**Author**: Roger D. Peng [rpeng@jhsph.edu](rpeng@jhsph.edu) with contributions from Duncan Murdoch and Barry Rowlingson; GPC library by Alan Murta

**Maintainer**: Roger D. Peng [rpeng@jhsph.edu](rpeng@jhsph.edu)

**License**: file LICENSE

**Depends**: R (<= 2.14.0), methods

**Imports**: graphics

**Date**: 2013-04-01

**URL**: [http://www.cs.man.ac.uk/~toby/gpc/](http://www.cs.man.ac.uk/~toby/gpc/), [http://github.com/rdpeng/gpclib](http://github.com/rdpeng/gpclib)

---

## `R` code

- Copy `R` code into the `R/` subdirectory
- There can be any number of files in this directory
- Usually separate out files into logical groups
- Code for all functions should be included here and not anywhere else in the package

---

## The `NAMESPACE` File

- Used to indicate which functions are **exported**
- Exported functions can be called by the user and are considered the public API
- Non-exported functions cannot be called directly by the user (but the code can be viewed)
- Hides implementation details from users and makes a cleaner package interface

---

## The `NAMESPACE` File

- You can also indicate what functions you **import** from other packages
- This allows your package to use other packages without making other packages visible to the user
- Importing a function loads the package but does not attach it to the search list

---

## The `NAMESPACE` File

Key directives

- `export("<function>")`
- `import("<package>")`
- `importFrom("<package>", "<function>")`

Also important

- `exportClasses("<class>")`
- `exportMethods("<generic>")`

---

## `NAMESPACE` File: `mvtsplot` package

```{r, eval=FALSE}
export("mvtsplot")
importFrom(graphics, "Axis")
import(splines)
```

---

## `NAMESPACE` File: `gpclib` package

```{r, eval=FALSE}
export("read.polyfile", "write.polyfile")
importFrom(graphics, plot)
exportClasses("gpc.poly", "gpc.poly.nohole")
exportMethods("show", "get.bbox", "plot", "intersect", "union", "setdiff",
              "[", "append.poly", "scale.poly", "area.poly", "get.pts",
              "coerce", "tristrip", "trangulate")
```

---

## Documentation

- Documentation files (`.Rd`) placed in `man/` subdirectory
- Written in a specific markup language
    - Another reason to limit exported functions
- You can document other things like concepts, package overview

---

## Help File Example: `line` Function

```
\name{line}
\alias{line}
\alias{residuals.tukeyline}
\title{Robust Line Fitting}
\description{
    Fit a line robustly as recommended in \emph{Exploratory Data Analysis}.
}
\usage{
line(x, y)
}
\arguments{
    \item{x, y}{the arguments can be any way of specifying x-y pairs. See \code{\link{xy.coords}}.}
}
\details{
    Cases with missing values are omitted.
    
    Long vectors are not supported.
}
\value{
    An object of class \code{"tukeyline"}.
    
    Methods are available for the generic functions \code{coef},
    \code{residuals}, \code{fitted}, and \code{print}.
}
\references{
    Tukey, J. W. (1977).
    \emph{Exploratory Data Analysis}
    Reading Massachusetts: Addison-Wesley.
}
```

---

## Building and Checking

- `R CMD build` is a command-line program that creates a package archive file (`.tar.gz`)
- `R CMD check` runs a battery of tests on the package
- You can run `R CMD build` or `R CMD check` from the command-line using a terminal or command-shell application
- You can also run them from `R` using the `system()` function

```{r, eval=FALSE}
system("R CMD build newpackage")
system("R CMD check newpackage")
```

---

## Checking

- `R CMD check` runs a battery of tests
- Documentation exists
- Code can be loaded, no major coding problems or errors
- Run examples in documentation
- Check docs match code
- All tests must pass to put package on CRAN

---

## Getting Started

- The `package.skeleton()` function in the `utils` package creates a "skeleton" `R` package
- Directory structure (`R/`, `man/`), `DESCRIPTION` file, `NAMESPACE` file, documentation files
- If there are functions visible in your workspace, it writes `R` code files to the `R\` directory
- Documentation stubs are created in `man\`
- You need to fill in the rest!

---

## Summary

- `R` packages provide a systematic way to make `R` code available to others
- Standards ensure that packages have a minimal amount of documentation and robustness
- Obtained from CRAN, Bioconductor, Github, etc.

---

## Summary

- Create a new directory with `R\` and `man\` subdirectories (or just use `package.skeleton()`)
- Write a `DESCRIPTION` file
- Copy `R` code into the `R\` subdirectory
- Write documentation files in `man\` subdirectory
- Write a `NAMESPACE` file with exports/imports
- Build and check